---
title: "Death by state in Peru"
author: "Emilio Sopprani, Grace Rico, Gonzalo Quintanilla, Mark Hegedus"
date: "2024-10-09"
output:
  html_document:
    code_folding: show
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE)
```

# - Helper functions
```{r}
show_head <- function(df) {
  head(df) %>% # passing down only few lines from the df
    knitr::kable(format = "html") %>% # showing the lines in html format
    kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% # little bit of style to avoid columns to be crowded
    kableExtra::scroll_box(width = "100%") # make it scrollable so the columns don't overflow the page
}

are_assumptions_met <- function(expected_values) {
  under_freq_of_5 <- 0 # init a counter variable, which counts those frequencies that are lower than 5
  for (i in expected_values) { # loop through all the expected values (frequencies)
    if (i < 1) { # if at least one frequency is lower than 1 we just return False
      return(F)
    } else if (i < 5) { # count frequencies that are less than 5
      under_freq_of_5 <- 1 + under_freq_of_5 # increase counter variable
    }
  }
  return(ifelse(under_freq_of_5 / length(expected_values) > 0.2, F, T)) # if frequencies of less than 5 have a higher proportion than 20% we return False, otherwise True
}
```


# - Import dataset
```{r}
fallecidos_covid <- read.csv("./fallecidos_covid.csv", sep = ";")

show_head(fallecidos_covid)
```

# - Wrangling dataset
```{r}
# Filtering our data to get rid of all the rubbish
filtered_data <- fallecidos_covid |> select(DEPARTAMENTO, SEXO) |> # Select only the variables we are going to use
  filter(DEPARTAMENTO != "MASCULINO") |> # There is a mislabeled state as "Masculino" So we are choosing all the entries but that one.
  drop_na() |> # Just in Case
  summarize(.by = c(DEPARTAMENTO, SEXO), n = n()) # we collapse all the Masculine and Femenine counts by departamento and include the number "n" of observations

# We are creating our contingency table
cont_table <- filtered_data |>
  pivot_wider(names_from = SEXO, values_from = n) |> # The sex needs to be separated into columns so it is pivoted wider.
  as.data.frame() # turned as a df

# Change the row names to each DEPARTAMENTO to get a nice Contingecy table

rownames(cont_table) <- cont_table$DEPARTAMENTO # Change the row names to the state instead of having a column with it

# change the names so all the non-spanish speaking peeps can understand
cont_table <- cont_table |>
  mutate(
    "male" = MASCULINO,
    "female" = FEMENINO
  )

# selecting the useful columns and transforming it as a matrix to put in chisq.test()

cont_table_matrix <- cont_table |>
  select(c(male, female)) |>
  as.matrix()

show_head(cont_table_matrix)
```

# - Hypotheses
- H~0~ = There is no difference in deaths by sex amongst each Peruvian state.
- H~a~ = There is a difference in deaths by sex amongst each Peruvian state.

Our chosen significance level is 𝛂 = 0.05.

# - Visualization
```{r}

#Proportional stacked bar plot - Does not show counts (hides data) but "cleaner" version of the mosaic plot
ggplot(filtered_data, aes(x = DEPARTAMENTO, y = n, fill = SEXO)) +
  geom_bar(stat = "identity", position = "fill") + #Scales height of bars so that they are proportions
  scale_y_continuous(labels = scales::percent) + #Converts numeric variables into percents
  labs(title = "Proportion of Male and Female COVID Deaths by State",
       x = "State", y = "Proportion", fill = "Sex") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) #Makes x-axis labs readable

```


# - Test statistics and degree of freedom
```{r}
x2 <- chisq.test(cont_table_matrix)
x2

# There is one group "LIMA METROPOLITANA" that has only 8 observations (4 male/4 female) but its fine bc we have 52 entries so no assumptions are violated :)
sprintf("Are assumptions met? %s", are_assumptions_met(x2$expected)) # using helper function to check whether the assumptions are met
```


# - P-value and conclusion
```{r}
sprintf("P-value: %s", x2$p.value)
```
- We conclude that there is a difference in deaths by sex amongst each Peruvian state, since our p-value is less than our significance level, which is 0.05.
